Equalizer 2.0\\Convergence of a Parallel Rendering Framework

### 2017-04-03 paperIEEE.tex
More FIXMEs address, typeset projects
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@42 db7c848b-7755-4df2-b763-8796797afa52

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
  \begin{abstract}
    Developing complex, real world graphics applications which utilize multiple GPUs and
    computers for interactive 3D rendering tasks can be a daunting task, in
    particular since it requires expertise in distributed systems and parallel
    rendering in addition to the application domain itself. We present a mature
    parallel rendering framework which provides a large set of features,
    algorithms and system integration for a wide range of real-world research
    and industry applications. Based on the basic architecture and
    implementation of the Equalizer parallel rendering framework presented in
    \cite{EMP:09}, we show how a variety of generic algorithms are integrated in the
    framework to help application development in many different domains, and we show
    a number of applications highlighting the diverse aspects and use cases of Equalizer.
  \end{abstract}

@@

%----------------------------------------------------------------------
\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
    untracked views running RTT DeltaGen for virtual car usability studies, and (c)
    Cave2 running a molecular visualization build based on Omegalib.}
  \label{FIG_teaser}
\end{figure*}

@@

Highlight integration of many features into a general framework vs one-off
research implementation of one algorithm. Also point out the various and very
diverse examples in Section 7, highlighting the variety of application and implementation
scenarios that can benefit from Equalizer.

The rest of this paper is structured as follows: First we introduce new related
work since the introduction of Equalizer. The main body of this paper presents
new performance features, virtual reality algorithms, usability features to
build complex applications, main novel features of the underlying Collage
network library and a quick overview of the main Equalizer-based applications. A
result section presents new experiments not previously published, followed by
the discussion and conclusion.

@@

In~\cite{EMP:09}, we presented the initial {\em Equalizer} parallel rendering framework,
which summarized the work in parallel rendering up to 2009. An extensive
Programming and User Guide provides in-depth documentation on using and
programming Equalizer~\cite{Eilemann:13}. In the following we assume these two
publications and their references as a starting point, and focus on the related
work published since then.

The concept of transparent OpenGL interception popularized by WireGL and
Chromium~\cite{HHNFAKK:02} has received little attention since 2009. While some
commercial implementations such as TechViz and MechDyne Conduit continue to
exist, on the research side only ClusterGL~\cite{NHM:11} has been presented.
ClusterGL employs the same approach as Chromium, but delivers a significantly
faster implementation of transparent OpenGL interception and distribution for
parallel rendering. CGLX~\cite{DK:11} tries to bring parallel execution
transparently to OpenGL applications, by emulating the GLUT API and intercepting
certain OpenGL calls. In contrast to frameworks like Chromium and ClusterGL
which distribute OpenGL calls, CGLX follows the distributed application
approach. This works transparently for trivial applications, but quickly
requires the application developer to address the complexities of a distributed
application, when mutable application state needs to be synchronized across
processes. For realistic applications, writing parallel applications remains the
only viable approach for scalable parallel rendering, as shown by the success of
Paraview, Visit and Equalizer-based applications.

On the other hand, software for driving and interacting with tiled display walls
has received significant attention, including Sage~\cite{Sage} and
Sage~2~\cite{Sage2} in particular. Sage was built entirely around the concept of
a shared framebuffer where all content windows are separate applications using
pixel streaming but is no longer actively supported. Sage 2 is a complete,
browser-centric reimplementation where each application is a web application
distributed across browser instances. DisplayCluster~\cite{DC}, and its
continuation Tide~\cite{tide}, also implement the shared framebuffer concept of
Sage, but provide a few native content applications integrated into the display
servers. These solutions implement a scalable display environment and are a
target display platform for scalable 3D graphics applications.

Equalizer itself has received significant attention within the research
community. Various algorithms to improve the parallel rendering performance have
been proposed: compression and region of interest during
compositing~\cite{MEP:10}; load-balancing resources for multi-display
installations~\cite{EEP:11}; asynchronous compositing and NUMA
optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}.
Additionally, complex large scale and out-of-core multiresolution
rendering approaches have been parallelized and implemented with
Equalizer~\cite{GMBP:10, GEMPG:13}, demonstrating the feasibility of
the framework to be used with complex rendering algorithms and 3D model
representations.

Furthermore, various applications and frameworks have used Equalizer for new research in
visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
visualizing simulations of neurocortical brain regions; Omegalib~\cite{Omegalib},
extending Equalizer for the hybrid 2D/3D environments such as the CAVE2;
visualization for remote sensing data~\cite{LK:09}, bino.

@@

The initial version of Equalizer implemented the basic sort-first (2D), sort-last (DB),
stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
present the newly added decomposition modes and motivate their use case. \fig{fig:compounds}
provides an overview of the new modes. We describe how to setup compound
descriptions in the framework in \cite{EMP:09}.

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
  \end{subfigure}\\[\medskipamount]
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source3"
    phase 2 period 3
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}\vspace{24ex}}
    \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source1"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source2"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source3"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
  inputframe { name "frame.source3" }
}
    \end{lstlisting}}
    \caption{\label{fig:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 3 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    pixel [ 1 0 3 1 ]
    outputframe {}
  }
  compound {
    channel "source2"
    pixel [ 2 0 3 1 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 3 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    subpixel [ 1 3 ]
    outputframe {}
  }
  compound {
    channel "source2"
    subpixel [ 1 3 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New Equalizer task decomposition modes and their compound descriptions for parallel rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to Equalizer
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined below in \sref{sec:threading}. The output frame
rate of the destination channel is smoothened using a {\em frame rate equalizer}
(\sref{sec:framerateEq}).

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks are queued and
processed by all source channels by polling a server-central queue. Prefetching
ensures that the task communication overlaps with rendering. As shown in~
\cite{SPEP:16}, these modes provide better performance due to being inherently
load-balanced, as long as there is an insignificant overhead for the render task
setup. This mode is transparent to Equalizer applications.

@@

Pixel compounds (\fig{fig:pixel}) decompose the destination channel by
interleaving rows or columns in image space. They are a variant of sort-first
decomposition which works well for fill-limited applications which are not
geometry bound. Source channels cannot reduce geometry load through view frustum
culling, since each source channel has almost the same frustum (only shifted by
some pixels), but applied to a reduced 2D viewport. However, the fragment load
on all source channels is very similar due to the interleaved distribution of
pixels. This functionality is transparent to Equalizer applications, and the
default compositing implementation uses the OpenGL stencil buffer to blit pixels
onto the destination channel.

@@

{\em Equalizers} are an addition to compound trees. They modify parameters of
their respective subtree at runtime to dynamically optimize the resource usage,
by each tuning one aspect of the decomposition. Due to their nature, they are
transparent to application developers, but might have application-accessible
parameters to tune their behavior. Resource equalization is the critical
component for scalable parallel rendering, and therefore the eponym for the
Equalizer project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
name for the framework as well as for this component, can't we change this?
eile: We can't change the product name, and changing the component name will be
hard. In fact, this is an artefact of the architecture: The runtime balancing of
resources is the core contribution of Equalizer, and therefore the equalizers
performing this task are the name-givers for the product. Modified paragraph
above for clarity.}

@@

The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
adapting the rendering resolution of a fill-limited application. In Equalizer,
this works by rendering into a source channel (typically on a FBO) separate to
the destination channel, and then scaling the rendering during the transfer
(typically through an on-GPU texture) to the destination channel. The DFR
equalizer monitors the rendering performance and accordingly adapts the
resolution of the source channel and zoom factor for the source to destination
transfer. If the performance and source channel resolutions allows, this will
not only subsample, but also supersample the destination channel to reduce
aliasing artefacts.

@@

Compression for the compositing step in parallel rendering is critical for
performance. This not only applies to the well-researched network transfer step,
but also for the transfer between GPU and CPU. Equalizer supports a variety of
compression algorithms, from very fast RLE encoding, YUV
subsampling on the GPU to JPEG compression. These algorithms are implemented as runtime-loaded
plugins, allowing easy extension and customization to application-specific
compression. In \cite{MEP:10}, we show this to be a critical step for
interactive performance at scale.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
large data scenarios. To easily customize the synchronization of all threads on
a single process, Equalizer implements three threading modes:  Full
synchronization, draw synchronization and asynchronous. Note that the execution
between nodes is always asynchronous, for up to \textsf{latency} frames.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. Equalizer has been used in many virtual reality
installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
at the KAUST visualization laboratory and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

Head tracking is the minimal feature needed to support immersive
installations. Equalizer does support multiple, independent tracked views
through the observer abstraction introduced in \sref{sec:observer}. Built-in
VRPN support enables the direct, application-transparent configuration of a VRPN
tracker device. Alternatively, an application can provide a $4\times 4$ tracking
matrix defining the transformation from the canvas coordinate system to the
observer. Both CAVE-like tracking with fixed projection surfaces and HMD
tracking modes are implemented.

@@

Traditional head tracking computes the left and right eye positions by using a
configurable interocular distance and the tracking matrix. However, human heads
are not symmetric, and by measuring individual users a more precise frustum can
be computed. Equalizer supports this through the optional configuration (file or
programmatically) of individual 3D positions for the left and right eye.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. \label{sec:swap} A canvas may have a software or hardware
swap barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a Collage barrier (\sref{sec:barrier}) to synchronize all segments, the
swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
the swap command. The hardware swap barrier is implemented using the
\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

On each canvas, the application can display one or more views. A view is a view
on a model, in the sense used by the MVC pattern. The view class is used by
Equalizer applications to define view-specific data for rendering, e.g., a
scene, viewing mode or camera. The application process manages this data, and
the render clients receive it for rendering.

@@

Switching a layout, as described above, or switching the stereo rendering mode,
may involve a different set of resources after the change, including the launch
and exit of render client processes. Equalizer solves this through a
reconfiguration step at the beginning of each rendering frame. Each resource
(channel, window, pipe, node) has an activation count, which is updated when the
layout or any other relevant rendering parameter is changed. When a resource is
found whose activation count does not match its current start/stopped state, the
resource is created or destroyed and \textsf{configInit} or \textsf{configExit}
are called accordingly. In the current implementation, a normal configuration
initialization or exit, as described in \cite{EMP:09}, uses the same code path
with all used resources transitioning to a running or stopped state,
accordingly. Since starting new resources typically requires object mapping and
associated data distribution, it is a costly operation. To improve performance
and robustness, Equalizer will flush the rendering pipeline whenever a runtime
reconfiguration is detected, before executing the change. This implementation
allows the late join or early exit, transparent to application developers.

@@

The discovery is implemented in a separate library, hwsd\FIXME{need to somehow identify this term in the text,
UPPERCASE or some special font type} (HardWare Service
Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
WGL windowing systems, as well as network interfaces on Linux, Mac OS X and
Windows. Furthermore, it detects the presence of VirtualGL to allow optimal
configuration of remote visualization clusters. The resources can be discovered
on the local workstation, and through the help of a simple daemon using the
zeroconf protocol, on a set of remote nodes within a visualization cluster. A
session identifier may be used to support multiple users on a single cluster.

The Equalizer server uses the hwsd library to discover local and remote
resources when an hwsd session name instead of a \textsf{.eqc} configuration
file is provided. A set of standard decomposition modes is configured, which can
be selected through activating the corresponding layout.

This versatile mechanism allows non-experts to configure asnd profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to Equalizer application developers.

@@

Qt is a popular window system with many application developers. Unfortunately,
it imposes a different threading model for window creation and event handling
compared to Equalizer. In Equalizer, each GPU rendering thread is independently
responsible for creating its windows, receiving the events and eventually
dispatching them to the application process main thread. This design is
motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may be moved to a different thread, and
events are signalled from the main thread.

@@

Collage provides networking functionality of different abstraction layers,
gradually providing higher level functionality for the programmer. The main
primitives in Collage are:

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
  \begin{abstract}
    Developing complex, real world graphics applications which utilize multiple GPUs and
    computers for interactive 3D rendering tasks can be a daunting task, in
    particular since it requires expertise in distributed systems and parallel
    rendering in addition to the application domain itself. We present a mature
    parallel rendering framework which provides a large set of features,
    algorithms and system integration for a wide range of real-world research
    and industry applications. Based on the basic architecture and
    implementation of the \textsf{Equalizer} parallel rendering framework presented in
    \cite{EMP:09}, we show how a variety of generic algorithms are integrated in the
    framework to help application development in many different domains, and we show
    a number of applications highlighting the diverse aspects and use cases of \textsf{Equalizer}.
  \end{abstract}

@@

%----------------------------------------------------------------------
\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Example \textsf{Equalizer} application use cases: (a) 192 Megapixel CAVE at
    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
    untracked views running RTT DeltaGen for virtual car usability studies, and (c)
    Cave2 running a molecular visualization build based on Omegalib.}
  \label{FIG_teaser}
\end{figure*}

@@

Highlight integration of many features into a general framework vs one-off
research implementation of one algorithm. Also point out the various and very
diverse examples in Section 7, highlighting the variety of application and implementation
scenarios that can benefit from \textsf{Equalizer}.

The rest of this paper is structured as follows: First we introduce new related
work since the introduction of \textsf{Equalizer}. The main body of this paper presents
new performance features, virtual reality algorithms, usability features to
build complex applications, main novel features of the underlying \textsf{Collage}
network library and a quick overview of the main \textsf{Equalizer}-based applications. A
result section presents new experiments not previously published, followed by
the discussion and conclusion.

@@

In~\cite{EMP:09}, we presented the initial \textsf{Equalizer} parallel rendering framework,
which summarized the work in parallel rendering up to 2009. An extensive
Programming and User Guide provides in-depth documentation on using and
programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following we assume these two
publications and their references as a starting point, and focus on the related
work published since then.

The concept of transparent OpenGL interception popularized by \textsf{WireGL}
and \textsf{Chromium}~\cite{HHNFAKK:02} has received little attention since
2009. While some commercial implementations such as \textsf{TechViz} and
\textsf{MechDyne Conduit} continue to exist, on the research side only
\textsf{ClusterGL}~\cite{NHM:11} has been presented. \textsf{ClusterGL} employs
the same approach as \textsf{Chromium}, but delivers a significantly faster
implementation of transparent OpenGL interception and distribution for parallel
rendering. \textsf{CGLX}~\cite{DK:11} tries to bring parallel execution
transparently to OpenGL applications, by emulating the GLUT API and intercepting
certain OpenGL calls. In contrast to frameworks like \textsf{Chromium} and
\textsf{ClusterGL} which distribute OpenGL calls, \textsf{CGLX} follows the
distributed application approach. This works transparently for trivial
applications, but quickly requires the application developer to address the
complexities of a distributed application, when mutable application state needs
to be synchronized across processes. For realistic applications, writing
parallel applications remains the only viable approach for scalable parallel
rendering, as shown by the success of \textsf{Paraview}, \textsf{Visit} and
\textsf{Equalizer}-based applications.

On the other hand, software for driving and interacting with tiled display walls
has received significant attention, including \textsf{Sage}~\cite{Sage} and
\textsf{Sage~2}~\cite{Sage2} in particular. \textsf{Sage} was built entirely
around the concept of a shared framebuffer where all content windows are
separate applications using pixel streaming but is no longer actively supported.
\textsf{Sage 2} is a complete, browser-centric reimplementation where each
application is a web application distributed across browser instances.
\textsf{DisplayCluster}~\cite{DC}, and its continuation
\textsf{Tide}~\cite{tide}, also implement the shared framebuffer concept of
\textsf{Sage}, but provide a few native content applications integrated into the
display servers. These solutions implement a scalable display environment and
are a target display platform for scalable 3D graphics applications.

\textsf{Equalizer} itself has received significant attention within the research
community. Various algorithms to improve the parallel rendering performance have
been proposed: compression and region of interest during
compositing~\cite{MEP:10}; load-balancing resources for multi-display
installations~\cite{EEP:11}; asynchronous compositing and NUMA
optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}.
Additionally, complex large scale and out-of-core multiresolution
rendering approaches have been parallelized and implemented with
Equalizer~\cite{GMBP:10, GEMPG:13}, demonstrating the feasibility of
the framework to be used with complex rendering algorithms and 3D model
representations.

Furthermore, various applications and frameworks have used \textsf{Equalizer}
for new research in visualization, such as \textsf{RTNeuron}~\cite{HBBES:13}, an
OpenGL renderer for visualizing simulations of neurocortical brain regions;

\textsf{Omegalib}~\cite{Omegalib}, extending \textsf{Equalizer} for the hybrid
2D/3D environments such as the CAVE2;

visualization for remote sensing data~\cite{LK:09}, bino.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first (2D), sort-last (DB),
stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
present the newly added decomposition modes and motivate their use case. \fig{fig:compounds}
provides an overview of the new modes. We describe how to setup compound
descriptions in the framework in \cite{EMP:09}.

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
  \end{subfigure}\\[\medskipamount]
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source3"
    phase 2 period 3
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}\vspace{24ex}}
    \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source1"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source2"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source3"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
  inputframe { name "frame.source3" }
}
    \end{lstlisting}}
    \caption{\label{fig:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 3 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    pixel [ 1 0 3 1 ]
    outputframe {}
  }
  compound {
    channel "source2"
    pixel [ 2 0 3 1 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 3 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    subpixel [ 1 3 ]
    outputframe {}
  }
  compound {
    channel "source2"
    subpixel [ 1 3 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New \textsf{Equalizer} task decomposition modes and their compound descriptions for parallel rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to \textsf{Equalizer}
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined below in \sref{sec:threading}. The output frame
rate of the destination channel is smoothened using a {\em frame rate equalizer}
(\sref{sec:framerateEq}).

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks are queued and
processed by all source channels by polling a server-central queue. Prefetching
ensures that the task communication overlaps with rendering. As shown in~
\cite{SPEP:16}, these modes provide better performance due to being inherently
load-balanced, as long as there is an insignificant overhead for the render task
setup. This mode is transparent to \textsf{Equalizer} applications.

@@

Pixel compounds (\fig{fig:pixel}) decompose the destination channel by
interleaving rows or columns in image space. They are a variant of sort-first
decomposition which works well for fill-limited applications which are not
geometry bound. Source channels cannot reduce geometry load through view frustum
culling, since each source channel has almost the same frustum (only shifted by
some pixels), but applied to a reduced 2D viewport. However, the fragment load
on all source channels is very similar due to the interleaved distribution of
pixels. This functionality is transparent to \textsf{Equalizer} applications, and the
default compositing implementation uses the OpenGL stencil buffer to blit pixels
onto the destination channel.

@@

{\em Equalizers} are an addition to compound trees. They modify parameters of
their respective subtree at runtime to dynamically optimize the resource usage,
by each tuning one aspect of the decomposition. Due to their nature, they are
transparent to application developers, but might have application-accessible
parameters to tune their behavior. Resource equalization is the critical
component for scalable parallel rendering, and therefore the eponym for the
\textsf{Equalizer} project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
name for the framework as well as for this component, can't we change this?
eile: We can't change the product name, and changing the component name will be
hard. In fact, this is an artefact of the sw architecture: The runtime balancing
of resources is the core contribution of Equalizer, and therefore the equalizers
performing this task are the name-givers for the product. Modified paragraph
above for clarity.}

@@

The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
adapting the rendering resolution of a fill-limited application. In \textsf{Equalizer},
this works by rendering into a source channel (typically on a FBO) separate to
the destination channel, and then scaling the rendering during the transfer
(typically through an on-GPU texture) to the destination channel. The DFR
equalizer monitors the rendering performance and accordingly adapts the
resolution of the source channel and zoom factor for the source to destination
transfer. If the performance and source channel resolutions allows, this will
not only subsample, but also supersample the destination channel to reduce
aliasing artefacts.

@@

Compression for the compositing step in parallel rendering is critical for
performance. This not only applies to the well-researched network transfer step,
but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a variety of
compression algorithms, from very fast RLE encoding, YUV
subsampling on the GPU to JPEG compression. These algorithms are implemented as runtime-loaded
plugins, allowing easy extension and customization to application-specific
compression. In \cite{MEP:10}, we show this to be a critical step for
interactive performance at scale.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
large data scenarios. To easily customize the synchronization of all threads on
a single process, \textsf{Equalizer} implements three threading modes:  Full
synchronization, draw synchronization and asynchronous. Note that the execution
between nodes is always asynchronous, for up to \textsf{latency} frames.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. \textsf{Equalizer} has been used in many virtual reality
installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
at the KAUST visualization laboratory and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

Head tracking is the minimal feature needed to support immersive
installations. \textsf{Equalizer} does support multiple, independent tracked views
through the observer abstraction introduced in \sref{sec:observer}. Built-in
VRPN support enables the direct, application-transparent configuration of a VRPN
tracker device. Alternatively, an application can provide a $4\times 4$ tracking
matrix defining the transformation from the canvas coordinate system to the
observer. Both CAVE-like tracking with fixed projection surfaces and HMD
tracking modes are implemented.

@@

Traditional head tracking computes the left and right eye positions by using a
configurable interocular distance and the tracking matrix. However, human heads
are not symmetric, and by measuring individual users a more precise frustum can
be computed. \textsf{Equalizer} supports this through the optional configuration (file or
programmatically) of individual 3D positions for the left and right eye.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. \label{sec:swap} A canvas may have a software or hardware
swap barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all segments, the
swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
the swap command. The hardware swap barrier is implemented using the
\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

On each canvas, the application can display one or more views. A view is a view
on a model, in the sense used by the MVC pattern. The view class is used by
\textsf{Equalizer} applications to define view-specific data for rendering, e.g., a
scene, viewing mode or camera. The application process manages this data, and
the render clients receive it for rendering.

@@

Switching a layout, as described above, or switching the stereo rendering mode,
may involve a different set of resources after the change, including the launch
and exit of render client processes. \textsf{Equalizer} solves this through a
reconfiguration step at the beginning of each rendering frame. Each resource
(channel, window, pipe, node) has an activation count, which is updated when the
layout or any other relevant rendering parameter is changed. When a resource is
found whose activation count does not match its current start/stopped state, the
resource is created or destroyed and \textsf{configInit} or \textsf{configExit}
are called accordingly. In the current implementation, a normal configuration
initialization or exit, as described in \cite{EMP:09}, uses the same code path
with all used resources transitioning to a running or stopped state,
accordingly. Since starting new resources typically requires object mapping and
associated data distribution, it is a costly operation. To improve performance
and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a runtime
reconfiguration is detected, before executing the change. This implementation
allows the late join or early exit, transparent to application developers.

@@

The discovery is implemented in a separate library, \textsf{hwsd} (HardWare Service
Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
WGL windowing systems, as well as network interfaces on Linux, Mac OS X and
Windows. Furthermore, it detects the presence of VirtualGL to allow optimal
configuration of remote visualization clusters. The resources can be discovered
on the local workstation, and through the help of a simple daemon using the
zeroconf protocol, on a set of remote nodes within a visualization cluster. A
session identifier may be used to support multiple users on a single cluster.

The \textsf{Equalizer} server uses the hwsd library to discover local and remote
resources when an hwsd session name instead of a \texttt{.eqc} configuration
file is provided. A set of standard decomposition modes is configured, which can
be selected through activating the corresponding layout.

This versatile mechanism allows non-experts to configure and profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to \textsf{Equalizer} application developers.

@@

Qt is a popular window system with many application developers. Unfortunately,
it imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread is independently
responsible for creating its windows, receiving the events and eventually
dispatching them to the application process main thread. This design is
motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may be moved to a different thread, and
events are signalled from the main thread.

@@

\textsf{Collage} provides networking functionality of different abstraction layers,
gradually providing higher level functionality for the programmer. The main
primitives in \textsf{Collage} are:
